---
title: "eheat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eheat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This package serves as a bridge between the ggplot2 and ComplexHeatmap packages.
Essentially, all ggplot2 geometries and operations can be utilized in
ComplexHeatmap through the `eheat` package, with the exception of facet
operations (and you shouldn't do it in `eheat` package). Fortunately,
ComplexHeatmap is capable of handling these operations independently, rendering
them unnecessary. 

## Installation

You can install the development version of `eheat` from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!requireNamespace("pak")) {
  install.packages("pak",
    repos = sprintf(
      "https://r-lib.github.io/p/pak/devel/%s/%s/%s",
      .Platform$pkgType, R.Version()$os, R.Version()$arch
    )
  )
}
pak::pkg_install("Yunuuuu/eheat")
```

```{r setup}
library(eheat)
```


Let's begin by creating some example data, following code was copied from
ComplexHeatmap book directly

```{r prepare_data}
set.seed(123)
nr1 <- 4
nr2 <- 8
nr3 <- 6
nr <- nr1 + nr2 + nr3
nc1 <- 6
nc2 <- 8
nc3 <- 10
nc <- nc1 + nc2 + nc3
mat <- cbind(
  rbind(
    matrix(rnorm(nr1 * nc1, mean = 1, sd = 0.5), nrow = nr1),
    matrix(rnorm(nr2 * nc1, mean = 0, sd = 0.5), nrow = nr2),
    matrix(rnorm(nr3 * nc1, mean = 0, sd = 0.5), nrow = nr3)
  ),
  rbind(
    matrix(rnorm(nr1 * nc2, mean = 0, sd = 0.5), nrow = nr1),
    matrix(rnorm(nr2 * nc2, mean = 1, sd = 0.5), nrow = nr2),
    matrix(rnorm(nr3 * nc2, mean = 0, sd = 0.5), nrow = nr3)
  ),
  rbind(
    matrix(rnorm(nr1 * nc3, mean = 0.5, sd = 0.5), nrow = nr1),
    matrix(rnorm(nr2 * nc3, mean = 0.5, sd = 0.5), nrow = nr2),
    matrix(rnorm(nr3 * nc3, mean = 1, sd = 0.5), nrow = nr3)
  )
)
mat <- mat[sample(nr, nr), sample(nc, nc)] # random shuffle rows and columns
rownames(mat) <- paste0("row", seq_len(nr))
colnames(mat) <- paste0("column", seq_len(nc))
small_mat <- mat[1:9, 1:9]
``` 

The central functions of the `eheat` package are `ggheat` and `gganno`. These
two functions encompass all the necessary functionalities. `ggheat` serves as a
substitute for the `ComplexHeatmap::Heatmap` function, while `gganno` replaces
all the `anno_*` functions within the ComplexHeatmap package, offering a
comprehensive solution for our requirements. One of the key advantages of using ggplot2 in ComplexHeatmap is the ease of plotting statistical annotations. Another benefit is that the legends can be internally extracted from the ggplot2 plot, eliminating the need for manual addition of legends.

## ggheat
Using `ggheat`, it is effortless to create a simple Heatmap. The default color
mapping was not consistent between ComplexHeatmap and ggplot2.

```{r heat_single}
draw(ggheat(small_mat))
```

You do not need to explicitly specify the color mapping as you can utilize the
`scale_*` function directly from ggplot2. All guide legends will directly
extracted from `ggplot2`. The essential parameter of `ggheat` is `ggfn`, which
accepts a ggplot2 object equipped with a default data and mappings established
via `ggplot(data, aes(.data$.column, .data$.row))`. By default, the data
includes 7 columns, each prefixed with a dot for caution.

* `.slice`: slice number, combine `.slice_row` and `.slice_column`.

* `.slice_row`: the slice row number.

* `.slice_column`: the slice column number.

* `.row` and `.column`: the row and column coordinates.

* `.row_index` and `.column_index`: the row and column index of the original

```{r heat_data}
pdf(NULL)
draw(ggheat(small_mat, function(x) {
  print(head(x$data))
  x
}))
dev.off()
```
  
The richness of the `scale_*` function in ggplot2 makes it easy to modify the color mapping according to our needs.
```{r}
draw(ggheat(small_mat, function(p) {
  # will use zero as midpoint
  p + scale_fill_gradient2()
}))
```

```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c(option = "magma")
}))
```

Legends can be controlled by `guide_*` function in ggplot2.
```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c(guide = guide_colorbar(direction = "horizontal"))
}))
```

You can add more geoms.
```{r}
draw(
  ggheat(small_mat, function(p) {
    p +
      geom_text(aes(label = sprintf("%d * %d", .row_index, .column_index)))
  })
)
```

You can also use the same way in ComplexHeatmap to prevent the internal rect
filling by setting `rect_gp = gpar(type = "none")`. The clustering is still
applied but nothing in drawn on the heatmap body. 
```{r}
draw(
  ggheat(small_mat, rect_gp = gpar(type = "none"))
)
```

Note that the background is different between ggplot2 and ComplexHeatmap.
However, the theme system in ggplot2 makes it easy to modify and customize the
background according to our preferences. 
```{r}
draw(
  ggheat(small_mat, function(p) {
    p +
      geom_text(aes(label = sprintf("%d * %d", .row_index, .column_index))) +
      theme_bw()
  }, rect_gp = gpar(type = "none"))
)
```

You can customize it easily use `geom_tile`.
```{r}
draw(
  ggheat(small_mat, function(p) {
    p +
      geom_tile(
        aes(fill = values),
        width = 1L, height = 1L,
        data = ~ dplyr::filter(p$data, .row <= .column)
      ) +
      geom_text(
        aes(label = sprintf("%d * %d", .row_index, .column_index)),
        data = ~ dplyr::filter(p$data, .row >= .column)
      ) +
      theme_bw()
  }, rect_gp = gpar(type = "none"))
)
```

All the functionalities of the `ComplexHeatmap::Heatmap` function can be used as is.
```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}, column_km = 2L))
```

```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}, column_km = 2L, row_km = 3))
```

We can combine `layer_fun` or `cell_fun` from ComplexHeatmap with `ggfn`
```{r}
draw(
  ggheat(small_mat,
    layer_fun = function(...) {
      grid.rect(gp = gpar(lwd = 2, fill = "transparent", col = "red"))
    }, column_km = 2L, row_km = 3
  )
)
```

The row names and column names should be controlled by the
`ComplexHeatmap::Heatmap` function, while the legends should be controlled by
`ggplot2`. However, the default legend name is taken from
`ComplexHeatmap::Heatmap` in order to maintain consistency. Nevertheless, you
can directly override it in `ggfn`.
```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}, column_km = 2L, row_km = 3, row_names_gp = gpar(col = "red")))
```

```{r}
draw(
  ggheat(small_mat, function(p) {
    p + scale_fill_viridis_c()
  },
  column_km = 2L, row_km = 3, row_names_gp = gpar(col = "red"),
  name = "ComplexHeatmap"
  )
)
```

```{r}
draw(
  ggheat(small_mat, function(p) {
    p + scale_fill_viridis_c(name = "ggplot2")
  },
  column_km = 2L, row_km = 3, row_names_gp = gpar(col = "red"),
  name = "ComplexHeatmap"
  )
)
```

## gganno
Both `gganno` and `gganno2` perform identical functions, but `gganno` is not
compatible with direct integration with `ComplexHeatmap::Heatmap`.  In such
cases, only an empty annotation region can be added. On the other hand,
`gganno2` can be seamlessly combined with both `ComplexHeatmap::Heatmap` and
`ggheat`, although legends will not be extracted.

The same with `ggheat`, the essential parameter of `gganno` is also `ggfn`,
which accepts a ggplot2 object equipped with a default data and mappings
established by `ggplot(data, aes(.data$.x (or .data$.y)))`. The original matrix
will be converted into a data.frame with another 3 columns added: 

* `.slice`: the slice row (which = "row") or column (which = "column")
  number. 

* `.x`/`.y`: indicating the x-axis (or y-axis) coordinates. Don't use
`ggplot2::coord_flip` to flip coordinates as it may disrupt internal
operations. 

* `.index`: denoting the row index of the original matrix, where rows are
  uniformly considered as observations and columns as variables. 
  

In general, we should just use `ggheat` and `gganno`.
```{r anno_point}
anno_data <- sample(1:10, nrow(small_mat))
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      # Note: vector will be converted one-column data.frame
      # with a column names `V1`
      matrix = anno_data,
      function(p) {
        p + geom_point(aes(.x, V1))
      }
    ), which = "column"
  )
))
```

Legends will also be extracted, in the similar manner like passing them into
`annotation_legend_list`. 
```{r anno_bar}
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + geom_bar(aes(y = V1, fill = factor(.index)), stat = "identity")
      }, height = unit(5, "cm")
    ), which = "column"
  )
), merge_legends = TRUE)
```

```{r anno_box}
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + geom_boxplot(aes(y = V1, fill = factor(.slice)))
      }, height = unit(5, "cm")
    ), which = "column"
  ), column_km = 2L
), merge_legends = TRUE)
```

```{r anno_violin}
box_matrix1 <- matrix(rnorm(ncol(small_mat)^2L, 10), nrow = ncol(small_mat))
colnames(box_matrix1) <- rep_len("group1", ncol(small_mat))
box_matrix2 <- matrix(rnorm(ncol(small_mat)^2L, 20), nrow = ncol(small_mat))
colnames(box_matrix2) <- rep_len("group2", ncol(small_mat))
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = cbind(box_matrix1, box_matrix2),
      function(p) {
        long_data <- ~ tidyr::pivot_longer(.x, starts_with("group"),
          names_to = "group"
        )
        p +
          geom_violin(
            aes(
              y = value, fill = factor(group),
              color = factor(.slice),
              group = paste(.slice, .index, group, sep = "-")
            ),
            data = long_data
          ) +
          geom_boxplot(
            aes(
              y = value, fill = factor(group),
              color = factor(.slice),
              group = paste(.slice, .index, group, sep = "-")
            ),
            width = 0.2,
            position = position_dodge(width = 0.9),
            data = long_data
          ) +
          scale_fill_brewer(
            name = "Group", type = "qual", palette = "Set3"
          ) +
          scale_color_brewer(
            name = "Slice", type = "qual", palette = "Set1"
          )
      }, height = unit(3, "cm")
    ), which = "column"
  ), column_km = 2L
), merge_legends = TRUE)
```

```{r, fig.height=8}
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + aes(y = V1) + geom_text(aes(label = .index))
      }, height = unit(2, "cm")
    ), which = "column"
  ),
  bottom_annotation = HeatmapAnnotation(
    foo = gganno(
      function(p) {
        p + aes(y = V1) +
          geom_text(aes(label = .index)) +
          scale_y_continuous(limits = rev)
      },
      matrix = anno_data,
      which = "column", height = unit(2, "cm")
    ),
    which = "column"
  ),
  right_annotation = HeatmapAnnotation(
    foo = gganno(
      function(p) {
        p + aes(x = V1) +
          geom_text(aes(label = .index))
      },
      matrix = anno_data,
      width = unit(3, "cm")
    ),
    which = "row"
  ),
  left_annotation = HeatmapAnnotation(
    foo = gganno(
      function(p) {
        p + aes(x = V1) +
          geom_text(aes(label = .index)) +
          scale_x_continuous(limits = rev)
      },
      matrix = anno_data,
      width = unit(3, "cm")
    ),
    which = "row"
  ),
  row_km = 2L, column_km = 2L,
), merge_legends = TRUE)
```

Finally, let's see the difference between `gganno2` and `gganno`.


`gganno2` will not extract the legend.
```{r}
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno2(
      matrix = anno_data,
      function(p) {
        p + geom_bar(aes(y = V1, fill = factor(.index)), stat = "identity")
      }
    ), which = "column"
  )
), merge_legends = TRUE)
```

But `gganno2` can work with `Heatmap` function.
```{r}
anno_data <- sample(1:10, nrow(small_mat))
draw(Heatmap(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno2(
      matrix = anno_data,
      function(p) {
        p + geom_point(aes(.x, V1))
      }
    ), which = "column"
  )
), merge_legends = TRUE)
```

`gganno` will just add a blank region in `Heatmap` function.
```{r}
draw(Heatmap(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + geom_bar(aes(y = V1, fill = factor(.index)), stat = "identity")
      }
    ), which = "column"
  )
), merge_legends = TRUE)
```
