---
title: "eheat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eheat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This package serves as a bridge between the ggplot2 and ComplexHeatmap packages.
Essentially, all ggplot2 geometries and operations can be utilized in
ComplexHeatmap through the `eheat` package, with the exception of facet
operations (and you shouldn't do it in `eheat` package). Fortunately,
ComplexHeatmap is capable of handling these operations independently, rendering
them unnecessary. 

## Installation

You can install the development version of `eheat` from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!requireNamespace("pak")) {
  install.packages("pak",
    repos = sprintf(
      "https://r-lib.github.io/p/pak/devel/%s/%s/%s",
      .Platform$pkgType, R.Version()$os, R.Version()$arch
    )
  )
}
pak::pkg_install("Yunuuuu/eheat")
```

```{r setup}
library(eheat)
library(ggplot2)
library(ComplexHeatmap)
```


Let's begin by creating some example data, following code was copied from
ComplexHeatmap book directly

```{r}
set.seed(123)
nr1 <- 4
nr2 <- 8
nr3 <- 6
nr <- nr1 + nr2 + nr3
nc1 <- 6
nc2 <- 8
nc3 <- 10
nc <- nc1 + nc2 + nc3
mat <- cbind(
  rbind(
    matrix(rnorm(nr1 * nc1, mean = 1, sd = 0.5), nr = nr1),
    matrix(rnorm(nr2 * nc1, mean = 0, sd = 0.5), nr = nr2),
    matrix(rnorm(nr3 * nc1, mean = 0, sd = 0.5), nr = nr3)
  ),
  rbind(
    matrix(rnorm(nr1 * nc2, mean = 0, sd = 0.5), nr = nr1),
    matrix(rnorm(nr2 * nc2, mean = 1, sd = 0.5), nr = nr2),
    matrix(rnorm(nr3 * nc2, mean = 0, sd = 0.5), nr = nr3)
  ),
  rbind(
    matrix(rnorm(nr1 * nc3, mean = 0.5, sd = 0.5), nr = nr1),
    matrix(rnorm(nr2 * nc3, mean = 0.5, sd = 0.5), nr = nr2),
    matrix(rnorm(nr3 * nc3, mean = 1, sd = 0.5), nr = nr3)
  )
)
mat <- mat[sample(nr, nr), sample(nc, nc)] # random shuffle rows and columns
rownames(mat) <- paste0("row", seq_len(nr))
colnames(mat) <- paste0("column", seq_len(nc))
small_mat <- mat[1:9, 1:9]
``` 

The central functions of the `eheat` package are `ggheat` and `gganno`. These two functions encompass all the necessary functionalities. `ggheat` serves as a substitute for the `ComplexHeatmap::Heatmap` function, while `gganno` replaces all the `anno_*` functions within the ComplexHeatmap package, offering a comprehensive solution for our requirements.

## ggheat
Using `ggheat`, it is effortless to create a simple Heatmap. 

```{r}
draw(ggheat(small_mat))
```

You do not need to explicitly specify the color mapping as you can utilize the
`scale_*` function directly from ggplot2. All guide legends will directly
extracted from `ggplot2`. The essential parameter of `ggheat` is `ggfn`, which
accepts a ggplot2 object equipped with a default data and mappings established
via `ggplot(data, aes(.data$.column, .data$.row))`. By default, the data
includes six columns, each prefixed with a dot for caution.

* `.slice_row`: the slice row number

* `.slice_column`: the slice column number

* `.row` and `.column`: the row and column coordinates

* `.row_index` and `.column_index`: the row and column index of the original
   matrix.
  
```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}))
```

You can add more geoms.
```{r}
draw(
  ggheat(small_mat, function(p) {
    p + geom_text(aes(label = sprintf("%d * %d", .row_index, .column_index)))
  })
)
```

All the functionalities of the `ComplexHeatmap::Heatmap` function can be used as is.
```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}, column_km = 2L))
```

```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}, column_km = 2L, row_km = 3))
```

The row names and column names should be controlled by the
`ComplexHeatmap::Heatmap` function, while the legends should be controlled by
`ggplot2`. However, the default legend name is taken from
`ComplexHeatmap::Heatmap` in order to maintain consistency. Nevertheless, you
can directly override it in `ggfn`.
```{r}
draw(ggheat(small_mat, function(p) {
  p + scale_fill_viridis_c()
}, column_km = 2L, row_km = 3, row_names_gp = gpar(col = "red")))
```

```{r}
draw(
  ggheat(small_mat, function(p) {
    p + scale_fill_viridis_c()
  },
  column_km = 2L, row_km = 3, row_names_gp = gpar(col = "red"),
  name = "ComplexHeatmap"
  )
)
```

```{r}
draw(
  ggheat(small_mat, function(p) {
    p + scale_fill_viridis_c(name = "ggplot2")
  },
  column_km = 2L, row_km = 3, row_names_gp = gpar(col = "red"),
  name = "ComplexHeatmap"
  )
)
```

## gganno
Both `gganno` and `gganno2` perform identical functions, but `gganno` is not
compatible with direct integration with `ComplexHeatmap::Heatmap`.  In such
cases, only an empty annotation region can be added. On the other hand,
`gganno2` can be seamlessly combined with both `ComplexHeatmap::Heatmap` and
`ggheat`, although legends will not be extracted.

The same with `ggheat`, the essential parameter of `gganno` is also `ggfn`,
which accepts a ggplot2 object equipped with a default data and mappings
established by `ggplot(data, aes(.data$x))`. The original matrix will be
converted into a data.frame with another 3 columns added: 

* `.slice`: the slice row (which = "row") or column (which = "column")
  number. 

* `.x`/`.y`: indicating the x-axis (or y-axis) coordinates. Don't use
`ggplot2::coord_flip` to flip coordinates as it may disrupt internal
operations. 

* `.index`: denoting the row index of the original matrix, where rows are
  uniformly considered as observations and columns as variables. 
  

In general, we should just use `ggheat` and `gganno`.
```{r}
anno_data <- sample(1:10, nrow(small_mat))
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      # Note: vector will be converted one-column data.frame
      # with a column names `V1`
      matrix = anno_data,
      function(p) {
        p + geom_point(aes(.x, V1))
      },
      which = "column"
    ), which = "column"
  )
))
```

Legends will also be extracted, in the similar manner like passing them into
`annotation_legend_list`. 
```{r}
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + geom_bar(aes(y = V1, fill = factor(.index)), stat = "identity")
      },
      which = "column", height = unit(5, "cm")
    ), which = "column"
  )
), merge_legends = TRUE)
```


```{r, fig.height=8}
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + aes(y = V1) + geom_text(aes(label = .index))
      },
      which = "column", height = unit(2, "cm")
    ), which = "column"
  ),
  bottom_annotation = HeatmapAnnotation(
    foo = gganno(
      function(p) {
        p + aes(y = V1) +
          geom_text(aes(label = .index)) +
          scale_y_reverse()
      },
      matrix = anno_data,
      which = "column", height = unit(2, "cm")
    ),
    which = "column"
  ),
  right_annotation = HeatmapAnnotation(
    foo = gganno(
      function(p) {
        p + aes(x = V1) +
          geom_text(aes(label = .index))
      },
      matrix = anno_data,
      which = "row", width = unit(3, "cm")
    ),
    which = "row"
  ),
  left_annotation = HeatmapAnnotation(
    foo = gganno(
      function(p) {
        p + aes(x = V1) +
          geom_text(aes(label = .index)) +
          scale_x_reverse()
      },
      matrix = anno_data,
      which = "row", width = unit(3, "cm")
    ),
    which = "row"
  ),
  row_km = 2L, column_km = 2L,
), merge_legends = TRUE)
```

Finally, let's see the difference between `gganno2` and `gganno`.


`gganno2` will not extract the legend.
```{r}
anno_data <- sample(1:10, nrow(small_mat))
draw(ggheat(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno2(
      matrix = anno_data,
      function(p) {
        p + geom_bar(aes(y = V1, fill = factor(.index)), stat = "identity")
      },
      which = "column"
    ), which = "column"
  )
), merge_legends = TRUE)
```

But `gganno2` can work with `Heatmap` function.
```{r}
anno_data <- sample(1:10, nrow(small_mat))
draw(Heatmap(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno2(
      matrix = anno_data,
      function(p) {
        p + geom_point(aes(.x, V1))
      },
      which = "column"
    ), which = "column"
  )
), merge_legends = TRUE)
```

`gganno` will just add a blank region in `Heatmap` function.
```{r}
draw(Heatmap(small_mat,
  top_annotation = HeatmapAnnotation(
    foo = gganno(
      matrix = anno_data,
      function(p) {
        p + geom_bar(aes(y = V1, fill = factor(.index)), stat = "identity")
      },
      which = "column"
    ), which = "column"
  )
), merge_legends = TRUE)
```
